---
title: "EU export and import to Russia and its neighbours"
execute:
  echo: false
format: 
    html:
        embed-resources: true
        code-fold: true
        df-print: paged
knitr: 
  opts_chunk: 
    echo: false
    message: false
    warning: false
---

# Introduction

Russia war on Ukraine had an impact to EU and Russia trade. The goal of this document is to analyze these changes.

The data was taken from [Eurostat](https://ec.europa.eu/eurostat/databrowser/view/DS-045409/legacyMultiFreq/table?lang=en). The [code](https://github.com/vzemlys/ruimpeks/eu_ru_export.qmd) and [data](https://github.com/vzemlys/ruimpeks/xlsx) can be found in the [Github repository](https://github.com/vzemlys/ruimpeks/). 

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)


read_eurostat <- function(xlsx_name, value_name) {
  texp <- read_xlsx(xlsx_name, sheet = 3, skip = 8)
  texp1 <- texp %>%
    filter(!(TIME...1 %in% c("FREQ (Labels)", "PARTNER (Labels)", "Special value", ":"))) %>%
    filter(!is.na(TIME...1)) %>%
    rename(partner = TIME...1, reporter = TIME...2) %>%
    pivot_longer(cols = -(partner:reporter), names_to = "month", values_to = value_name) %>%
    filter(grepl("-", month)) %>%
    mutate(month = ymd(paste(month, "01", sep = "-"))) %>%
    arrange(partner, reporter, month) %>%
    mutate(across(.cols = all_of(value_name), .fns = as.numeric))
  texp1
}
fixNA <- function(x) {
  x[is.na(x)] <- 0
  x
}

cn <- read.csv("meta/salys.csv") %>% select(partner = country, country = name)
eu_cn <- read.csv("meta/eu_countries.csv")
eu_texp <- read_eurostat("xlsx/eu_total_export.xlsx", value_name = "total_export")
eu_timp <- read_eurostat("xlsx/eu_total_import.xlsx", value_name = "total_import")
eu_eexp <- read_eurostat("xlsx/eu_energy_export.xlsx", value_name = "energy_export")
eu_eimp <- read_eurostat("xlsx/eu_energy_import.xlsx", value_name = "energy_import")

eu_tr0 <- eu_texp %>%
  left_join(eu_timp) %>%
  left_join(eu_eexp) %>%
  left_join(eu_eimp) %>%
  mutate(across(.cols = c("total_export", "total_import", "energy_export", "energy_import"), .fns = fixNA)) %>%
  mutate(
    export = total_export - energy_export,
    import = total_import - energy_import
  ) %>%
  left_join(cn) %>%
  inner_join(eu_cn)

eu_tr <- filter(eu_tr0) %>% filter(month <= "2022-10-01")
```
# EU trade  with Russia and neighbouring countries 

## EU export to Russia and eighbouring countries

```{r}
cols <- RColorBrewer::brewer.pal(7, "Set1")[c(5, 3, 6, 2, 4, 1, 7)]
eu_tr1 <- eu_tr %>%
  group_by(country, month) %>%
  summarise(export = sum(export), import = sum(import)) %>%
  ungroup()

ggplot(aes(x = month, y = export, fill = country), data = eu_tr1) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "EU export to Russia and neighbouring countries")
```

## EU export to Russia neighbours

```{r}
ggplot(aes(x = month, y = export, fill = country), data = eu_tr1 %>% filter(country != "Russia")) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols[-6]) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "EU export to Russia neighbours")
```



## EU import from Russia and neighbouring countries

```{r}
cols <- RColorBrewer::brewer.pal(7, "Set1")[c(5, 3, 6, 2, 4, 1, 7)]
eu_tr1 <- eu_tr %>%
  group_by(country, month) %>%
  summarise(export = sum(export), import = sum(import)) %>%
  ungroup()

ggplot(aes(x = month, y = import, fill = country), data = eu_tr1) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "EU import from Russia and neighbouring countries")
```

## EU import from Russia neighbours

```{r}
ggplot(aes(x = month, y = import, fill = country), data = eu_tr1 %>% filter(country != "Russia")) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols[-6]) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "EU import from Russia neighbours")
```


# EU trade after war growth compared to previous year

```{r}
eu_trf <- eu_tr %>%
  mutate(
    period =
      ifelse(month >= "2021-03-01" & month <= "2021-10-01", 2021,
        ifelse(month >= "2022-03-01" & month <= "2022-10-01", 2022, NA)
      )
  ) %>%
  filter(!is.na(period)) %>%
  group_by(partner = country, reporter = eu_country, period) %>%
  summarise(export = sum(export), import = sum(import))

eu_erf <- eu_trf %>%
  filter(partner == "Russia") %>%
  arrange(reporter, period) %>%
  mutate(egrowth = round(100 * (export / lag(export) - 1), 2), igrowth = round(100 * (import / lag(import) - 1), 2))

eu_erf1 <- eu_erf %>%
  filter(period == 2022) %>%
  arrange(reporter) %>%
  mutate(epos = egrowth >= 0, ipos = igrowth >= 0) %>%
  mutate(reporter = factor(reporter, levels = sort(eu_cn$eu_country, decreasing = TRUE)))

eu_enb <- eu_trf %>%
  filter(partner != "Russia") %>%
  group_by(reporter, period) %>%
  summarise(export = sum(export), import = sum(import)) %>%
  mutate(egrowth = round(100 * (export / lag(export) - 1), 2), igrowth = round(100 * (import / lag(import) - 1), 2))

eu_enb1 <- eu_enb %>%
  filter(period == 2022) %>%
  arrange(reporter) %>%
  mutate(reporter = factor(reporter, levels = sort(eu_cn$eu_country, decreasing = TRUE))) %>%
  mutate(epos = egrowth >= 0, ipos = igrowth >= 0)
```

## EU export growth to Russia

```{r}
ggplot(aes(x = reporter, y = egrowth, fill = epos), data = eu_erf1) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols[c(4, 6)], guide = "none") +
  labs(y = "Export growth, %", x = "", title = "EU export to Russia, 2022 after war growth compared to 2021")
```

## EU export growth to Russia neighbours

```{r}
ggplot(aes(x = reporter, y = egrowth, fill = epos), data = eu_enb1) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols[c(4, 6)], guide = "none") +
  labs(y = "Export growth, %", x = "", title = "EU export to Russia neighbours, 2022 after war growth compared to 2021")
```


## EU import growth from Russia

```{r}
ggplot(aes(x = reporter, y = igrowth, fill = ipos), data = eu_erf1) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols[c(4, 6)], guide = "none") +
  labs(y = "Export growth, %", x = "", title = "EU import from Russia, 2022 after war growth compared to 2021")
```

## EU import growth from Russia neighbours

```{r}
ggplot(aes(x = reporter, y = igrowth, fill = ipos), data = eu_enb1) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols[c(4, 6)], guide = "none") +
  labs(y = "Export growth, %", x = "", title = "EU import from Russia neighbours, 2022 after war growth compared to 2021")
```

# Sanctions

```{r}
eu_es <- readxl::read_xlsx("xlsx/eu_export_sanctioned.xlsx") %>%
  mutate(month = ymd(month))

es1 <- eu_es %>%
  inner_join(cn) %>%
  inner_join(eu_cn)

es2 <- es1 %>%
  group_by(country, month, code_length) %>%
  summarise(sanctioned_export = sum(sanctioned_export)) %>%
  ungroup()
```

```{r}
ggplot(aes(x = month, y = sanctioned_export, fill = country), data = es2 %>% filter(code_length == 4)) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Sanctioned EU export to Russia and neighbouring countries, 4 digits code")
```

```{r}
ggplot(aes(x = month, y = sanctioned_export, fill = country), data = es2 %>% filter(country != "Russia", code_length == 4)) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols[-6]) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Sanctioned EU export to Russia neighbours, 4 digits code")
```

```{r}
ggplot(aes(x = month, y = sanctioned_export, fill = country), data = es2 %>% filter(code_length == 8)) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Sanctioned EU export to Russia and neighbouring countries, 8 digits code")
```

```{r}
ggplot(aes(x = month, y = sanctioned_export, fill = country), data = es2 %>% filter(country != "Russia", code_length == 8)) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols[-6]) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Sanctioned EU export to Russia neighbours, 8 digits code")
```
