---
title: "On EU trade with Russia and its neighbours after Russian invasion of Ukraine"
execute:
  echo: false
format: 
    html:
        embed-resources: true
        code-fold: true
        df-print: paged
        toc: true
        toc-location: left
knitr: 
  opts_chunk: 
    echo: false
    message: false
    warning: false
---



```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(stringr)
library(arrow)
library(DT)

source("R/read_eurostat.R")

cn <- read.csv("meta/salys.csv") %>% select(partner = country, country = name)
eu_cn <- read.csv("meta/eu_countries.csv")

eu_texp <- read_eurostat("xlsx/eu_total_trade_4630244_spreadsheet.xlsx", flow = TRUE)

eu_texp1 <- eu_texp %>%
  mutate(name = paste(flow, code, sep = "_")) %>%
  pivot_wider(id_cols = c("reporter", "partner", "month"), values_from = value, names_from = "name", values_fill = 0) %>%
  rename(
    total_export = export_TOTAL, total_import = import_TOTAL,
    energy_import = import_27, energy_export = export_27
  ) %>%
  mutate(
    export = total_export - energy_export,
    import = total_import - energy_import
  ) %>%
  left_join(cn) %>%
  inner_join(eu_cn)


totals <- eu_texp1 %>%
  group_by(month) %>%
  summarise(across(.cols = c("energy_import", "energy_export", "total_import", "total_export", "export", "import"), sum)) %>%
  ungroup() %>%
  mutate(pi = round(100 * energy_import / total_import, 2), pe = round(100 * energy_export / total_export, 2))

cmp22 <- c(ymd("2022-03-01"), ymd("2022-11-01"))
cmp21 <- cmp22 - years(1)


eu_tr <- eu_texp1 %>% filter(month <= cmp22[2])

eu_trf <- eu_tr %>%
  mutate(
    period =
      ifelse(month >= cmp21[1] & month <= cmp21[2], 2021,
        ifelse(month >= cmp22[1] & month <= cmp22[2], 2022, NA)
      )
  ) %>%
  filter(!is.na(period)) %>%
  group_by(partner = country, reporter = eu_country, period) %>%
  summarise(export = sum(export), import = sum(import))

eu_totals <- eu_trf %>%
  mutate(russia = partner == "Russia") %>%
  group_by(russia, period) %>%
  summarise(export = sum(export), import = sum(import)) %>%
  ungroup() %>%
  arrange(russia, period) %>%
  group_by(russia) %>%
  mutate(egrowth = round(100 * (export / lag(export) - 1), 2), igrowth = round(100 * (import / lag(import) - 1), 2)) %>%
  ungroup()

ru_data <- eu_totals %>% filter(period == 2022, russia)
nb_data <- eu_totals %>% filter(period == 2022, !russia)
```
# Introduction

Russia's war on Ukraine had an impact to EU and Russia trade. The goal of this document is to give a bird's eye view of these changes using the EU trade data from Eurostat. 

## Main findings

1. Export to Russia dropped by `r round(abs(ru_data$egrowth[1]))`% after Russia invasion of Ukraine on February 24th, 2022.  Export to Russia neighbours increased by  `r round(nb_data$egrowth[1])`%.
3. Sanctioned goods export to Russia fell by 71%. Sanctioned good export to Russia neighbours increased by 95%.
4. More that 100 sanctioned export markets have grown by more than 100% after the war has started. This constitutes 21% of total sanctioned exports. This strongly suggests that EU countries are evading sanctions by exporting to Russia neighbours instead.

# EU overall trade with Russia and neighbouring countries 

Energy import constitutes about 70% of total imports. It distorts overall picture so it was excluded from the analysis. Energy exports constitute around 1% of total exports, but their calculation is complicated for countries connected to BRELL grid. For this reason it is excluded too.


In the post-invasion  period EU countries exports to Russia fell by `r round(abs(ru_data$egrowth[1]))`% to `r round(ru_data$export[1]/1e9)` billion euro compared to similar period in 2021. However EU countries export to Russia neighbours rose by `r round(nb_data$egrowth[1])`% to `r round(nb_data$export[1]/1e9)` billion euro. 

In the post-invasion war period EU countries imports to Russia fell by `r round(abs(ru_data$igrowth[1]))`% to `r round(ru_data$import[1]/1e9)` billion euro compared to similar period in 2021. EU countries import from Russia neighbours rose by `r round(nb_data$igrowth[1])`% to `r round(nb_data$import[1]/1e9)` billion euro. 

## Overall trade 

### EU exports to Russia and neighbouring countries

```{r}
eu_tr1 <- eu_tr %>%
  group_by(country, month) %>%
  summarise(export = sum(export), import = sum(import)) %>%
  ungroup()

eu_trt <- eu_tr1 %>%
  group_by(country) %>%
  summarise(export = sum(export)) %>%
  arrange(export)
clevs <- eu_trt$country

eu_tr1 <- eu_tr1 %>% mutate(country = factor(country, levels = clevs))
cols <- RColorBrewer::brewer.pal(7, "Set1")[c(4, 5, 6, 7, 3, 2, 1)]
cols_true <- RColorBrewer::brewer.pal(7, "Set1")

ggplot(aes(x = month, y = export, fill = country), data = eu_tr1) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "EU export to Russia and neighbouring countries")
```

### EU export to Russia neighbours

```{r}
ggplot(aes(x = month, y = export, fill = country), data = eu_tr1 %>% filter(country != "Russia")) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols[-8]) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "EU export to Russia neighbours")
```

### EU import from Russia and neighbouring countries

```{r}

ggplot(aes(x = month, y = import, fill = country), data = eu_tr1) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "EU import from Russia and neighbouring countries")
```

### EU import from Russia neighbours

```{r}
ggplot(aes(x = month, y = import, fill = country), data = eu_tr1 %>% filter(country != "Russia")) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols[-8]) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "EU import from Russia neighbours")
```


## Overall trade growth

Comparison is made between the months following the Russian invasion and corresponding months in 2021. 
```{r}

eu_erf <- eu_trf %>%
  filter(partner == "Russia") %>%
  arrange(reporter, period) %>%
  mutate(egrowth = round(100 * (export / lag(export) - 1), 2), igrowth = round(100 * (import / lag(import) - 1), 2))

eu_erf1 <- eu_erf %>%
  filter(period == 2022) %>%
  arrange(reporter) %>%
  mutate(epos = egrowth >= 0, ipos = igrowth >= 0) %>%
  mutate(reporter = factor(reporter, levels = sort(eu_cn$eu_country, decreasing = TRUE)))

eu_enb <- eu_trf %>%
  filter(partner != "Russia") %>%
  group_by(reporter, period) %>%
  summarise(export = sum(export), import = sum(import)) %>%
  mutate(egrowth = round(100 * (export / lag(export) - 1), 2), igrowth = round(100 * (import / lag(import) - 1), 2))

eu_enb1 <- eu_enb %>%
  filter(period == 2022) %>%
  arrange(reporter) %>%
  mutate(reporter = factor(reporter, levels = sort(eu_cn$eu_country, decreasing = TRUE))) %>%
  mutate(epos = egrowth >= 0, ipos = igrowth >= 0)
```

### EU export growth to Russia

```{r}
eu_erf2 <- eu_erf1 %>% arrange(egrowth)
eu_erf3 <- eu_erf2 %>% mutate(reporter = factor(as.character(reporter), levels = eu_erf2$reporter))

ggplot(aes(x = reporter, y = egrowth, fill = epos), data = eu_erf3) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols_true[c(2, 1)], guide = "none") +
  labs(y = "Export growth, %", x = "", title = "EU export growth to Russia after the start of the war")
```

### EU export growth to Russia neighbours

```{r}
eu_enb2 <- eu_enb1 %>% arrange(egrowth)
eu_enb3 <- eu_enb2 %>% mutate(reporter = factor(as.character(reporter), levels = eu_enb2$reporter))

ggplot(aes(x = reporter, y = egrowth, fill = epos), data = eu_enb3) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols_true[c(2, 1)], guide = "none") +
  labs(y = "Export growth, %", x = "", title = "EU export growth to Russia neighbours after the start of the war")
```

### EU import growth from Russia

```{r}
eu_erf4 <- eu_erf1 %>% arrange(igrowth)
eu_erf5 <- eu_erf4 %>% mutate(reporter = factor(as.character(reporter), levels = eu_erf4$reporter))


ggplot(aes(x = reporter, y = igrowth, fill = ipos), data = eu_erf5) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols_true[c(2, 1)], guide = "none") +
  labs(y = "Import growth, %", x = "", title = "EU import growth from Russia after the start of the war")
```

### EU import growth from Russia neighbours

```{r}
eu_enb4 <- eu_enb1 %>% arrange(igrowth)
eu_enb5 <- eu_enb4 %>% mutate(reporter = factor(as.character(reporter), levels = eu_enb4$reporter))

ggplot(aes(x = reporter, y = igrowth, fill = ipos), data = eu_enb5) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols_true[c(2, 1)], guide = "none") +
  labs(y = "Import growth, %", x = "", title = "EU import growth from Russia neighbours after the start of the war")
```

# Effect of sanctions on export

Sanctions put on exports are listed by their CN code in the following [document](https://eur-lex.europa.eu/legal-content/EN/TXT/?uri=CELEX%3A02014R0833-20221007). See @sec-appendix for technical details how the data was collected. The sanctions mentioned both CN 4 digits and CN 8 digits code. Only CN 8 digit codes were used for analysis.

There are quotas and exceptions for certain sanctioned goods, which allow export, so it is expected that sanctioned export is non zero. The effect of quotas and exceptions is not taken to account.

## Export of sanctioned goods 

### To Russia with its neighbours

```{r}
espq <- read_parquet("raw/eu_export_sanctions_data.parquet")
eu_es <- espq %>%
  group_by(partner, reporter, month) %>%
  summarise(sanctioned_export = sum(value))

es1 <- eu_es %>%
  inner_join(cn) %>%
  inner_join(eu_cn)

es2 <- es1 %>%
  group_by(country, month) %>%
  summarise(sanctioned_export = sum(sanctioned_export)) %>%
  ungroup() %>%
  mutate(country = factor(country, levels = clevs))
```



```{r}

ggplot(aes(x = month, y = sanctioned_export, fill = country), data = es2) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Sanctioned EU export to Russia and neighbouring countries")
```

### To Russia neighbours
```{r}
ggplot(aes(x = month, y = sanctioned_export, fill = country), data = es2 %>% filter(country != "Russia")) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols[-8]) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Sanctioned EU export to Russia neighbours")
```
## Sanctioned export growth


```{r}
eu_strf <- es1 %>%
  mutate(
    period =
      ifelse(month >= cmp21[1] & month <= cmp21[2], 2021,
        ifelse(month >= cmp22[1] & month <= cmp22[2], 2022, NA)
      )
  ) %>%
  filter(!is.na(period)) %>%
  group_by(partner = country, reporter = eu_country, period) %>%
  summarise(sanctioned_export = sum(sanctioned_export))

eu_serf <- eu_strf %>%
  filter(partner == "Russia") %>%
  arrange(reporter, period) %>%
  mutate(egrowth = round(100 * (sanctioned_export / lag(sanctioned_export) - 1), 2))

eu_serf1 <- eu_serf %>%
  filter(period == 2022) %>%
  arrange(reporter) %>%
  mutate(epos = egrowth >= 0) %>%
  mutate(reporter = factor(reporter, levels = sort(eu_cn$eu_country, decreasing = TRUE)))

eu_senb <- eu_strf %>%
  filter(partner != "Russia") %>%
  group_by(reporter, period) %>%
  summarise(sanctioned_export = sum(sanctioned_export)) %>%
  mutate(egrowth = round(100 * (sanctioned_export / lag(sanctioned_export) - 1), 2))

eu_senb1 <- eu_senb %>%
  filter(period == 2022) %>%
  arrange(reporter) %>%
  mutate(reporter = factor(reporter, levels = sort(eu_cn$eu_country, decreasing = TRUE))) %>%
  mutate(epos = egrowth >= 0)
```

### EU sanctioned export growth to Russia

```{r}
eu_serf2 <- eu_serf1 %>% arrange(egrowth)
eu_serf3 <- eu_serf2 %>% mutate(reporter = factor(as.character(reporter), levels = eu_serf2$reporter))

ggplot(aes(x = reporter, y = egrowth, fill = epos), data = eu_serf3) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols_true[c(2, 1)], guide = "none") +
  labs(y = "Export growth, %", x = "", title = "EU sanctioned export growth to Russia after the start of the war")
```

### EU sanctioned export growth to Russia neighbours

Malta sanctioned export grew by 4156% so it is not shown on the graph. 

```{r}
eu_senb2 <- eu_senb1 %>% arrange(egrowth)
eu_senb3 <- eu_senb2 %>% mutate(reporter = factor(as.character(reporter), levels = eu_senb2$reporter))

ggplot(aes(x = reporter, y = egrowth, fill = epos), data = eu_senb3) +
  geom_col() +
  ylim(c(min(eu_senb3$egrowth), 500)) +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols_true[c(2, 1)], guide = "none") +
  labs(y = "Export growth (capped at 500%), %", x = "", title = "EU sanctioned export growth to Russia neighbours after the start of the war")
```

```{r}
scd <- espq %>%
  count(code, product) %>%
  select(-n) %>%
  mutate(head = substring(code, 1, 4))

espq1 <- espq %>%
  mutate(
    period =
      ifelse(month >= cmp21[1] & month <= cmp21[2], 2021,
        ifelse(month >= cmp22[1] & month <= cmp22[2], 2022, NA)
      )
  ) %>%
  inner_join(cn) %>%
  inner_join(eu_cn) %>%
  filter(!is.na(period)) %>%
  group_by(partner = country, reporter = eu_country, period, code, product) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(partner, reporter, code, product, period) %>%
  group_by(partner, reporter, code, product) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2)) %>%
  ungroup()

espq2 <- espq1 %>%
  mutate(head = substring(code, 1, 4)) %>%
  group_by(partner, reporter, period, head) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(partner, reporter, head, period) %>%
  group_by(partner, reporter, head) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2))

espq3 <- espq1 %>%
  group_by(partner, period, code, product) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(partner, code, product, period) %>%
  group_by(partner, code, product) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2)) %>%
  ungroup()

espq4 <- espq1 %>%
  group_by(reporter, period, code, product) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(reporter, code, product, period) %>%
  group_by(reporter, code, product) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2)) %>%
  ungroup()

espq5 <- espq1 %>%
  group_by(reporter, period) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(reporter, period) %>%
  group_by(reporter) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2)) %>%
  ungroup()

total_se <- espq1 %>%
  mutate(russia = partner == "Russia") %>%
  group_by(russia, period) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(russia, period) %>%
  group_by(russia) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2)) %>%
  ungroup()
```

### EU sanctioned export 

Absolute values Russian and its neighbours combined.

```{r}
es_set <- espq5 %>%
  filter(period == 2022) %>%
  arrange(value)
es_set1 <- es_set %>% mutate(reporter = factor(reporter, levels = es_set$reporter))

ggplot(aes(x = reporter, y = value), data = es_set1 %>% mutate(value = value / 1e6)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(y = "Export, millions €", x = "", title = "EU sanctioned export, after start of war")
```


```{r}
es_seg <- espq5 %>%
  filter(period == 2022) %>%
  arrange(egrowth) %>%
  mutate(epos = egrowth > 0)
es_seg1 <- es_seg %>% mutate(reporter = factor(reporter, levels = es_seg$reporter))

ggplot(aes(x = reporter, y = egrowth, fill = epos), data = es_seg1 %>% mutate(value = value / 1e6)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  scale_fill_manual(values = cols_true[c(2, 1)], guide = "none") +
  labs(y = "Export, growth %", x = "", title = "EU sanctioned export growth, after start of war")

tb <- espq1 %>%
  filter(period == 2022) %>%
  filter(value > 10e6 & egrowth > 100) %>%
  mutate(growth = egrowth / 100) %>%
  arrange(-value) %>%
  select(reporter, partner, code, product, value, growth)

tb1 <- espq1 %>%
  filter(period == 2022) %>%
  mutate(growth = egrowth / 100) %>%
  arrange(-value) %>%
  select(reporter, partner, code, product, value, growth)
```

## Sanction anomalies

Export to Russia sanction evasion should be visible by increased export to other countries. Define anomaly as the export which is more than 10 million euro in value and which doubled after a war.

### Anomalous markets

Define by market an EU country, its export destination and good defined by 8 digit CN classifier. 

Using the anomaly definition `r nrow(tb)`  export categories stand out. It is interesting that there are `r nrow(tb %>% filter(partner == "Russia"))`` categories of sanctioned exports to Russia which experienced triple digit growth. 

The biggest of anomalous exports in term of value are motor cars. Lithuania topping the list with 173 millions euro export to Belarus which tripled compared to previous year. Germany exports of motor cars to Belarus and Russia grew ~8 times for cars with cylinder capacity between 1.5 and 2 litres. Germany exports of motor cars to Belarus with cylinder capacity larger than 2 litres grew 26 times. These Germany exports amounted to 422 milion euro of exports after the war.

Poland exported 139 millions of euro worth turbojets to Kazakhstan, and these exports grew 3 times. Czechia exported 233 millions euro of smartphones to Kazakhstan, Armenia and Georgia after the war, which is notable as in 2021 there were no such exports at the same period. 

Latvia exported 43 millions euro worth of sparling wine to Russia and these exports grew 4 times. 

Greece exported 75 millions euro worth of jet fuel to Georgia. These exports grew by ~4500 times compared to similar period in 2021.


```{r}


datatable(tb,
  options =
    list(
      dom = "lpt",
      pageLength = 10,
      lengthMenu = c(10, 50, nrow(tb)),
      columnDefs = list(list(
        targets = "_all",
        render = JS(
          "function(data, type, row, meta) {",
          "return type === 'display' && data != null && data.length > 30 ?",
          "'<span title=\"' + data + '\">' + data.substr(0, 30) + '...</span>' : data;",
          "}"
        )
      ))
    ),
  class = "display", rownames = FALSE, filter = "top"
) %>%
  formatCurrency(c("value"), currency = "€") %>%
  formatPercentage(c("growth"), 2)
``` 

The growth with empty values in the table means that the export in year 2021 from March to November was zero. The full name of exported goods appears if you hover on truncated name.

The full table with all the markets, their after the start of the war export value and growth from the same period in year 2021 can be found [here](https://github.com/vzemlys/ruimpeks/tree/main/xlsx/eu_sanctioned_export_growth.xlsx). Note the file size is 11MB.

# Appendix {#sec-appendix}

## Data

The data used comes from Eurostat dataset on international trade in [goods](https://ec.europa.eu/eurostat/databrowser/view/DS-045409/legacyMultiFreq/table?lang=en). 

All the code and data resides in the [Github repository](https://github.com/vzemlys/ruimpeks/):

1. [Code for this document](https://github.com/vzemlys/ruimpeks/blob/main/eu_ru_trade.qmd).  
2. [Total trade data](https://github.com/vzemlys/ruimpeks/tree/main/xlsx). 
3. [CN codes for export sanctions](https://github.com/vzemlys/ruimpeks/blob/main/raw/input_codes.csv).
4. [Sanctioned export data](https://github.com/vzemlys/ruimpeks/tree/main/raw) and the [code to prepare it](https://github.com/vzemlys/ruimpeks/blob/main/R/prepare_data.R).
5. [Sanctioned exports CN 8 digits codes](https://github.com/vzemlys/ruimpeks/blob/main/meta/export_sanctions_codes.csv).

## Citation

To cite this document use the following: V. Zemlys-Balevičius, "On EU trade with Russia and its neighbours after Russian invasion of Ukraine", vzemlys.github.io/ruimpeks, 25 January, 2022, https://vzemlys.github.io/ruimpeks/eu_ru_trade.html
