---
title: "Lithuania trade with Russia"
execute:
  echo: false
format: 
    html:
        embed-resources: true
        code-fold: true
        df-print: paged
        toc: true
        toc-location: left
knitr: 
  opts_chunk: 
    echo: false
    message: false
    warning: false
---



```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(stringr)
library(arrow)
library(DT)

source("R/read_eurostat.R")

cn <- read.csv("meta/salys.csv") %>% select(partner = country, country = name)
eu_cn <- read.csv("meta/eu_countries.csv")

eu_texp <- read_eurostat("xlsx/eu_total_trade_4630244_spreadsheet.xlsx", flow = TRUE)

eu_texp1 <- eu_texp %>%
  mutate(name = paste(flow, code, sep = "_")) %>%
  pivot_wider(id_cols = c("reporter", "partner", "month"), values_from = value, names_from = "name", values_fill = 0) %>%
  rename(
    total_export = export_TOTAL, total_import = import_TOTAL,
    energy_import = import_27, energy_export = export_27
  ) %>%
  mutate(
    export = total_export - energy_export,
    import = total_import - energy_import
  ) %>%
  left_join(cn) %>%
  inner_join(eu_cn) %>%
  filter(reporter == "Lithuania")


totals <- eu_texp1 %>%
  group_by(month) %>%
  summarise(across(.cols = c("energy_import", "energy_export", "total_import", "total_export", "export", "import"), sum)) %>%
  ungroup() %>%
  mutate(pi = round(100 * energy_import / total_import, 2), pe = round(100 * energy_export / total_export, 2))

cmp22 <- c(ymd("2022-03-01"), ymd("2022-11-01"))
cmp21 <- cmp22 - years(1)


eu_tr <- eu_texp1 %>% filter(month <= cmp22[2])

eu_trf <- eu_tr %>%
  mutate(
    period =
      ifelse(month >= cmp21[1] & month <= cmp21[2], 2021,
        ifelse(month >= cmp22[1] & month <= cmp22[2], 2022, NA)
      )
  ) %>%
  filter(!is.na(period)) %>%
  group_by(partner = country, reporter = eu_country, period) %>%
  summarise(export = sum(export), import = sum(import))

eu_totals <- eu_trf %>%
  mutate(russia = partner == "Russia") %>%
  group_by(russia, period) %>%
  summarise(export = sum(export), import = sum(import)) %>%
  ungroup() %>%
  arrange(russia, period) %>%
  group_by(russia) %>%
  mutate(egrowth = round(100 * (export / lag(export) - 1), 2), igrowth = round(100 * (import / lag(import) - 1), 2)) %>%
  ungroup()

ru_data <- eu_totals %>% filter(period == 2022, russia)
nb_data <- eu_totals %>% filter(period == 2022, !russia)

te21 <- eu_totals %>%
  filter(period == 2021) %>%
  .$export %>%
  sum()
te22 <- eu_totals %>%
  filter(period == 2022) %>%
  .$export %>%
  sum()

g <- round(100 * (te22 / te21 - 1), 2)
```
# Introduction

This document gives an overview of Lithuania trade with Russia and its neighbouring countries.

Export to Russia dropped by `r round(abs(ru_data$egrowth[1]))`% after Russia invasion of Ukraine on February 24th, 2022.  Export to Russia neighbours increased by  `r round(nb_data$egrowth[1])`%.

Total export to Russia and neighbouring countries increased by `r g`% from `r round(te21/1e9,2)` billion euros to `r round(te22/1e9,2)` billion euros.



## Overall trade 

### Lithuania exports to Russia and neighbouring countries

```{r}
eu_tr1 <- eu_tr %>%
  group_by(country, month) %>%
  summarise(export = sum(export), import = sum(import)) %>%
  ungroup()

eu_trt <- eu_tr1 %>%
  group_by(country) %>%
  summarise(export = sum(export)) %>%
  arrange(export)
clevs <- eu_trt$country

eu_tr1 <- eu_tr1 %>% mutate(country = factor(country, levels = clevs))
cols <- RColorBrewer::brewer.pal(7, "Set1")[c(4, 5, 6, 7, 3, 2, 1)]
cols_true <- RColorBrewer::brewer.pal(7, "Set1")

ggplot(aes(x = month, y = export, fill = country), data = eu_tr1) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Lithuania export to Russia and neighbouring countries")
```

### Lithuania export to Russia neighbours

```{r}
ggplot(aes(x = month, y = export, fill = country), data = eu_tr1 %>% filter(country != "Russia")) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols[-8]) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Lithuania export to Russia neighbours")
```

### Lithuania import from Russia and neighbouring countries

```{r}

ggplot(aes(x = month, y = import, fill = country), data = eu_tr1) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Lithuania import from Russia and neighbouring countries")
```

### Lithiania import from Russia neighbours

```{r}
ggplot(aes(x = month, y = import, fill = country), data = eu_tr1 %>% filter(country != "Russia")) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols[-8]) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Lithuania import from Russia neighbours")
```

# Effect of sanctions on export

Sanctions put on exports are listed by their CN code in the following [document](https://eur-lex.europa.eu/legal-content/EN/TXT/?uri=CELEX%3A02014R0833-20221007). See @sec-appendix for technical details how the data was collected. The sanctions mentioned both CN 4 digits and CN 8 digits code. Only CN 8 digit codes were used for analysis.

There are quotas and exceptions for certain sanctioned goods, which allow export, so it is expected that sanctioned export is non zero. The effect of quotas and exceptions is not taken to account.

## Export of sanctioned goods 

### To Russia with its neighbours

```{r}
espq <- read_parquet("raw/eu_export_sanctions_data.parquet")
eu_es <- espq %>%
  group_by(partner, reporter, month) %>%
  summarise(sanctioned_export = sum(value))

es1 <- eu_es %>%
  inner_join(cn) %>%
  inner_join(eu_cn) %>%
  filter(eu_country == "Lithuania")

es2 <- es1 %>%
  group_by(country, month) %>%
  summarise(sanctioned_export = sum(sanctioned_export)) %>%
  ungroup() %>%
  mutate(country = factor(country, levels = clevs))
```



```{r}

ggplot(aes(x = month, y = sanctioned_export, fill = country), data = es2) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Sanctioned LT export to Russia and neighbouring countries")
```

### To Russia neighbours
```{r}
ggplot(aes(x = month, y = sanctioned_export, fill = country), data = es2 %>% filter(country != "Russia")) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(name = "Country", values = cols[-8]) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€")) +
  labs(x = "Month", y = "", title = "Sanctioned Lithuania export to Russia neighbours")
```
```{r}
scd <- espq %>%
  count(code, product) %>%
  select(-n) %>%
  mutate(head = substring(code, 1, 4))

espq1 <- espq %>%
  mutate(
    period =
      ifelse(month >= cmp21[1] & month <= cmp21[2], 2021,
        ifelse(month >= cmp22[1] & month <= cmp22[2], 2022, NA)
      )
  ) %>%
  inner_join(cn) %>%
  inner_join(eu_cn) %>%
  filter(!is.na(period)) %>%
  group_by(partner = country, reporter = eu_country, period, code, product) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(partner, reporter, code, product, period) %>%
  group_by(partner, reporter, code, product) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2)) %>%
  ungroup() %>%
  filter(reporter == "Lithuania")

espq2 <- espq1 %>%
  mutate(head = substring(code, 1, 4)) %>%
  group_by(partner, reporter, period, head) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(partner, reporter, head, period) %>%
  group_by(partner, reporter, head) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2))

espq3 <- espq1 %>%
  group_by(partner, period, code, product) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(partner, code, product, period) %>%
  group_by(partner, code, product) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2)) %>%
  ungroup()

espq4 <- espq1 %>%
  group_by(reporter, period, code, product) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(reporter, code, product, period) %>%
  group_by(reporter, code, product) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2)) %>%
  ungroup()

espq5 <- espq1 %>%
  group_by(reporter, period) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(reporter, period) %>%
  group_by(reporter) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2)) %>%
  ungroup()

total_se <- espq1 %>%
  mutate(russia = partner == "Russia") %>%
  group_by(russia, period) %>%
  summarise(value = sum(value)) %>%
  ungroup() %>%
  arrange(russia, period) %>%
  group_by(russia) %>%
  mutate(egrowth = round(100 * (value / lag(value) - 1), 2)) %>%
  ungroup()
```


```{r}

tb <- espq1 %>%
  filter(period == 2022) %>%
  filter(value > 1e6 & egrowth > 100) %>%
  mutate(growth = egrowth / 100) %>%
  arrange(-value) %>%
  select(reporter, partner, code, product, value, growth)

tb1 <- espq1 %>%
  filter(period == 2022) %>%
  mutate(growth = egrowth / 100) %>%
  arrange(-value) %>%
  select(reporter, partner, code, product, value, growth)
```

## Sanction anomalies


### Anomalous markets

Define by market an EU country, its export destination and good defined by 8 digit CN classifier. 

Using the anomaly definition `r nrow(tb)`  export categories stand out. It is interesting that there are `r nrow(tb %>% filter(partner == "Russia"))`` categories of sanctioned exports to Russia which experienced triple digit growth. 

```{r}


datatable(tb,
  options =
    list(
      dom = "lpt",
      pageLength = 10,
      lengthMenu = c(10, 50, 100, nrow(tb)),
      columnDefs = list(list(
        targets = "_all",
        render = JS(
          "function(data, type, row, meta) {",
          "return type === 'display' && data != null && data.length > 30 ?",
          "'<span title=\"' + data + '\">' + data.substr(0, 30) + '...</span>' : data;",
          "}"
        )
      ))
    ),
  class = "display", rownames = FALSE, filter = "top"
) %>%
  formatCurrency(c("value"), currency = "€") %>%
  formatPercentage(c("growth"), 2)
``` 

The growth with empty values in the table means that the export in year 2021 from March to November was zero. The full name of exported goods appears if you hover on truncated name.

The full table with all the markets, their after the start of the war export value and growth from the same period in year 2021 can be found [here](https://github.com/vzemlys/ruimpeks/tree/main/xlsx/eu_sanctioned_export_growth.xlsx). Note the file size is 11MB.

# Appendix {#sec-appendix}

## Data

The data used comes from Eurostat dataset on international trade in [goods](https://ec.europa.eu/eurostat/databrowser/view/DS-045409/legacyMultiFreq/table?lang=en). 

All the code and data resides in the [Github repository](https://github.com/vzemlys/ruimpeks/):

1. [Code for this document](https://github.com/vzemlys/ruimpeks/blob/main/eu_ru_trade.qmd).  
2. [Total trade data](https://github.com/vzemlys/ruimpeks/tree/main/xlsx). 
3. [CN codes for export sanctions](https://github.com/vzemlys/ruimpeks/blob/main/raw/input_codes.csv).
4. [Sanctioned export data](https://github.com/vzemlys/ruimpeks/tree/main/raw) and the [code to prepare it](https://github.com/vzemlys/ruimpeks/blob/main/R/prepare_data.R).
5. [Sanctioned exports CN 8 digits codes](https://github.com/vzemlys/ruimpeks/blob/main/meta/export_sanctions_codes.csv).

## Citation

To cite this document use the following: V. Zemlys-Balevičius, "On EU trade with Russia and its neighbours after Russian invasion of Ukraine", vzemlys.github.io/ruimpeks, 25 January, 2022, https://vzemlys.github.io/ruimpeks/eu_ru_trade.html
